---
title: "ABALOBI Kenya Fisheries Monitoring"
subtitle: "Chale Beach Management Unit (BMU) Analysis"
author: "Lorenzo Longobardi (<a href='mailto:l.longobardi@cgiar.org'>l.longobardi@cgiar.org</a>)"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    smooth-scroll: true
    link-external-newwindow: true
    citations-hover: true
    footnotes-hover: true
    fig-width: 10
    fig-height: 6
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  echo: false
editor: visual
---

```{css}
.ai-disclaimer {
  background: linear-gradient(135deg, #dc3545 0%, #c82333 100%);
  color: white;
  padding: 0.5rem 1rem;
  margin: 0.5rem 0 1rem 0;
  border-radius: 0.25rem;
  border-left: 3px solid #a71e2a;
  box-shadow: 0 1px 3px rgba(220, 53, 69, 0.15);
  font-size: 0.85rem;
}

.ai-disclaimer .ai-title {
  font-weight: 600;
  font-size: 0.9rem;
  margin-bottom: 0.25rem;
  display: flex;
  align-items: center;
}

.ai-disclaimer .ai-title::before {
  content: "⚠️";
  margin-right: 0.4rem;
  font-size: 0.9rem;
}

.ai-disclaimer .ai-text {
  font-size: 0.8rem;
  line-height: 1.3;
  margin: 0;
  opacity: 0.95;
}

.card {
  border: 1px solid #dee2e6;
  border-radius: 0.375rem;
  padding: 1rem;
  margin-bottom: 1rem;
  background: #f8f9fa;
}

.metric-card {
  text-align: center;
  padding: 1.5rem;
  background: linear-gradient(135deg, #667eea 0%, #764ba2 100%);
  color: white;
  border-radius: 0.5rem;
  margin: 0.5rem;
}

.metric-number {
  font-size: 2.5rem;
  font-weight: bold;
  display: block;
}

.metric-label {
  font-size: 0.9rem;
  opacity: 0.9;
}

.highlight-box {
  background: #e3f2fd;
  border-left: 4px solid #2196f3;
  padding: 1rem;
  margin: 1rem 0;
}

.chale-banner {
  background: linear-gradient(135deg, #2E8B57 0%, #20B2AA 100%);
  color: white;
  padding: 1.5rem;
  border-radius: 0.5rem;
  text-align: center;
  margin: 1rem 0;
}
```

::::: ai-disclaimer
::: ai-title
AI-Generated Content Notice
:::

::: ai-text
Parts of this report were generated using AI assistance and have been human reviewed and validated. All data analysis, interpretations, and conclusions have been verified for accuracy and relevance.
:::
:::::

::: chale-banner
# 🏝️ CHALE BMU

## Multi-Site Fisheries Monitoring Analysis
:::

## Executive Summary {.highlight-box}

This report provides a comprehensive analysis of fishing activity across the **Chale Beach Management Unit (BMU)**, encompassing three key landing sites: **Chale-Jeza**, **Chale Island**, and **Kinondo Ngwani**. Together, these sites represent the majority of monitored fishing activity in the region, providing critical insights into local fishing patterns, species composition, and management needs across this important coastal fishing area.

## Data Overview

```{r setup}
#| include: false
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(DT)
library(knitr)
library(scales)
library(viridis)
library(patchwork)
library(apexcharter)
library(highcharter)
library(tidyr)

# Custom theme
theme_modern <- function() {
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", margin = margin(b = 20)),
    plot.subtitle = element_text(size = 12, color = "gray60", margin = margin(b = 20)),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(size = 11, face = "bold")
  )
}

# Set default theme
theme_set(theme_modern())
```

```{r data-load}
# Load and process data
abdat <- read_xlsx("data/ABALOBI Monitor_WCS All Communities_20-07-25.xlsx") |>
  mutate(
    tripstart = as_datetime(as.numeric(tripStartTimestamp) / 1000),
    tripend = as_datetime(as.numeric(tripEndTimestamp) / 1000),
    trip_duration_hrs = as.numeric((tripend - tripstart) / 3600),
    fishername = paste(fisherFirstName, fisherLastName, sep = "_"),
    TripLoggedDay = str_extract(TripLoggedDay, "[0-9]{4}-[0-9]{2}-[0-9]{2}"),
    TripLoggedDay = as_date(TripLoggedDay),
    trip_month = month(tripstart, label = TRUE),
    trip_hour = hour(tripstart),
    trip_day = wday(tripstart, label = TRUE)
  )

# Filter for Chale BMU (all 3 sites) and create processed dataset
chale_bmu <- abdat |>
  filter(monitoringSite_name_eng %in% c("Chale-Jeza", "Chale Island", "Kinondo Ngwani")) |>
  select(
    trip_id,
    landing_site_en = monitoringSite_name_eng,
    landing_site_loc = fishingSites_name_local,
    landing_date = TripLoggedDay,
    fisher_name = fishername,
    gender = fisherGender,
    trip_duration_hrs,
    trip_type,
    boat_type = isBoatType,
    boat_crew = boatCrewTotal,
    harvesterGroupTotal,
    isGroup,
    catch_taxon = catchList_name_eng,
    catch_basket_weight = catchBasketWeight,
    catch_weight = catchList_totalWeight,
    tripstart,
    trip_month,
    trip_hour,
    trip_day,
    gear_type = `Gear type`
  ) |>
  mutate(
    catch_weight = as.numeric(catch_weight),
    catch_basket_weight = as.numeric(catch_basket_weight)
  )

# Calculate key metrics for Chale BMU
total_records <- nrow(chale_bmu)
unique_trips <- length(unique(chale_bmu$trip_id))
unique_fishers <- length(unique(chale_bmu$fisher_name))
unique_sites <- length(unique(chale_bmu$landing_site_en))
date_range <- range(chale_bmu$landing_date, na.rm = TRUE)
unique_species <- length(unique(chale_bmu$catch_taxon[!is.na(chale_bmu$catch_taxon)]))
```

::::::::: card
### Chale BMU Key Metrics

:::::::: {style="display: flex; flex-wrap: wrap; justify-content: space-around;"}
::: metric-card
[`r comma(total_records)`]{.metric-number} [Total Records]{.metric-label}
:::

::: metric-card
[`r comma(unique_trips)`]{.metric-number} [Unique Trips]{.metric-label}
:::

::: metric-card
[`r comma(unique_fishers)`]{.metric-number} [Active Fishers]{.metric-label}
:::

::: metric-card
[`r unique_species`]{.metric-number} [Species Recorded]{.metric-label}
:::

::: metric-card
[`r unique_sites`]{.metric-number} [Landing Sites]{.metric-label}
:::
::::::::
:::::::::

**Data Collection Period:** `r date_range[1]` to `r date_range[2]`

## Landing Site Comparison within Chale BMU

```{r site-analysis}
# Landing site breakdown analysis
site_summary <- chale_bmu |>
  filter(!is.na(landing_site_en)) |>
  group_by(landing_site_en) |>
  summarise(
    total_records = n(),
    unique_trips = n_distinct(trip_id),
    unique_fishers = n_distinct(fisher_name),
    total_catch_weight = sum(catch_weight, na.rm = TRUE),
    avg_catch_weight = mean(catch_weight, na.rm = TRUE),
    species_count = n_distinct(catch_taxon, na.rm = TRUE),
    shore_trips = sum(trip_type == "shore", na.rm = TRUE),
    boat_trips = sum(trip_type == "boat", na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    percentage_records = round(total_records / sum(total_records) * 100, 1),
    shore_percentage = round(shore_trips / (shore_trips + boat_trips) * 100, 1)
  ) |>
  arrange(desc(total_records))

# Site comparison visualization
site_summary |>
  apex(
    type = "column",
    mapping = aes(x = landing_site_en, y = total_records)
  ) |>
  ax_title(
    text = "Fishing Activity by Landing Site within Chale BMU",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors("#2E8B57") |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_xaxis(title = list(text = "Landing Site")) |>
  ax_yaxis(title = list(text = "Number of Records"))

# Catch performance by site
# site_summary |>
#   apex(
#     type = "scatter",
#     mapping = aes(x = unique_trips, y = avg_catch_weight, size = total_catch_weight)
#   ) |>
#   ax_title(
#     text = "Site Performance: Trip Volume vs Average Catch",
#     subtitle = list(text = "Bubble size represents total catch weight"),
#     style = list(fontSize = "16px", fontWeight = "bold")
#   ) |>
#   ax_colors("#20B2AA") |>
#   ax_chart(toolbar = list(show = TRUE)) |>
#   ax_xaxis(title = list(text = "Number of Unique Trips")) |>
#   ax_yaxis(title = list(text = "Average Catch Weight (kg)"))

# Site method preferences
site_method_data <- chale_bmu |>
  filter(!is.na(landing_site_en) & !is.na(trip_type)) |>
  group_by(landing_site_en, trip_type) |>
  summarise(trips = n_distinct(trip_id), .groups = "drop")

site_method_data |>
  apex(
    type = "bar",
    mapping = aes(x = landing_site_en, y = trips, fill = trip_type)
  ) |>
  ax_title(
    text = "Fishing Method Preferences by Landing Site",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors(c("#FF5722", "#2196F3")) |>
  ax_chart(toolbar = list(show = TRUE), stacked = TRUE) |>
  ax_xaxis(title = list(text = "Landing Site")) |>
  ax_yaxis(title = list(text = "Number of Trips")) |>
  ax_legend(position = "top")
```

```{r site-table}
# Site comparison summary table
site_summary |>
  select(-shore_trips, -boat_trips) |>
  mutate(
    avg_catch_weight = round(avg_catch_weight, 2),
    total_catch_weight = round(total_catch_weight, 1)
  ) |>
  DT::datatable(
    caption = "Landing Site Comparison within Chale BMU",
    options = list(pageLength = 5, scrollX = TRUE),
    rownames = FALSE
  ) |>
  DT::formatRound(c("avg_catch_weight", "total_catch_weight"), 2) |>
  DT::formatStyle(
    "percentage_records",
    background = styleColorBar(site_summary$percentage_records, "#E8F4FD"),
    backgroundSize = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Fisher Demographics & Participation

```{r demographics}
# Gender analysis
gender_summary <- chale_bmu |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(
    records = n(),
    unique_fishers = n_distinct(fisher_name),
    percentage = round(n() / nrow(filter(chale_bmu, !is.na(gender))) * 100, 1),
    .groups = "drop"
  )

# Top active fishers
top_fishers <- chale_bmu |>
  count(fisher_name, sort = TRUE) |>
  slice_head(n = 15) |>
  mutate(fisher_name = str_replace(fisher_name, "_", " "))

# Gender distribution pie chart
gender_summary |>
  apex(
    type = "donut",
    mapping = aes(x = gender, y = records)
  ) |>
  ax_title(
    text = "Fisher Gender Distribution",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors(c("#E91E63", "#2196F3", "#FF9800", "#9C27B0")) |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_legend(position = "bottom")

```

```{r demographics-table}
# Gender summary table
gender_summary |>
  DT::datatable(
    caption = "Fisher Gender Distribution Summary",
    options = list(pageLength = 5, scrollX = TRUE),
    rownames = FALSE
  ) |>
  DT::formatStyle(
    "percentage",
    background = styleColorBar(gender_summary$percentage, "#E8F5E8"),
    backgroundSize = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Temporal Fishing Patterns

```{r temporal-analysis}
# Monthly patterns
monthly_summary <- chale_bmu |>
  filter(!is.na(trip_month)) |>
  group_by(trip_month) |>
  summarise(
    trips = n_distinct(trip_id),
    records = n(),
    avg_catch = mean(catch_weight, na.rm = TRUE),
    .groups = "drop"
  )

# Hourly patterns  
hourly_summary <- chale_bmu |>
  filter(!is.na(trip_hour)) |>
  group_by(trip_hour) |>
  summarise(
    trips = n_distinct(trip_id),
    records = n(),
    .groups = "drop"
  )

# Weekly patterns
weekly_summary <- chale_bmu |>
  filter(!is.na(trip_day)) |>
  group_by(trip_day) |>
  summarise(
    trips = n_distinct(trip_id),
    records = n(),
    .groups = "drop"
  )

# Monthly seasonal pattern
monthly_summary |>
  apex(
    type = "column",
    mapping = aes(x = trip_month, y = trips)
  ) |>
  ax_title(
    text = "Seasonal Fishing Patterns in Chale BMU",
    subtitle = list(text = "Number of fishing trips by month"),
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors("#FF6B35") |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_plotOptions(
    column = list(
      borderRadius = 4,
      columnWidth = "70%"
    )
  ) |>
  ax_xaxis(title = list(text = "Month")) |>
  ax_yaxis(title = list(text = "Number of Trips"))

# Hourly activity pattern with area chart
hourly_clean <- hourly_summary |> filter(trips > 5)
highchart() |>
  hc_add_series(
    data = hourly_clean,
    type = "area",
    hcaes(x = trip_hour, y = trips),
    name = "Trips",
    color = "#00BCD4",
    fillOpacity = 0.7
  ) |>
  hc_title(
    text = "Daily Departure Times in Chale BMU",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  hc_subtitle(text = "Trip departure times throughout the day") |>
  hc_xAxis(
    title = list(text = "Hour of Day"),
    tickInterval = 2,
    min = 0,
    max = 23
  ) |>
  hc_yAxis(title = list(text = "Number of Trips")) |>
  hc_chart(zoomType = "x") |>
  hc_exporting(enabled = TRUE)

# Weekly pattern
weekly_summary |>
  apex(
    type = "column",
    mapping = aes(x = trip_day, y = trips)
  ) |>
  ax_title(
    text = "Weekly Fishing Activity Patterns",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors("#4CAF50") |>
  ax_chart(toolbar = list(show = TRUE))
```

## Fishing Methods & Techniques

```{r fishing-methods}
# Trip type analysis
trip_type_summary <- chale_bmu |>
  filter(!is.na(trip_type)) |>
  group_by(trip_type) |>
  summarise(
    records = n(),
    trips = n_distinct(trip_id),
    percentage = round(n() / nrow(filter(chale_bmu, !is.na(trip_type))) * 100, 1),
    avg_duration = mean(trip_duration_hrs, na.rm = TRUE),
    avg_catch = mean(catch_weight, na.rm = TRUE),
    .groups = "drop"
  )

# Boat crew analysis
crew_summary <- chale_bmu |>
  filter(!is.na(boat_crew) & boat_crew > 0 & boat_crew < 15) |>
  group_by(boat_crew) |>
  summarise(
    records = n(),
    avg_catch = mean(catch_weight, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(records >= 5)

# Trip type distribution
trip_type_summary |>
  apex(
    type = "donut",
    mapping = aes(x = trip_type, y = records)
  ) |>
  ax_title(
    text = "Fishing Method Distribution",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors(c("#FF5722", "#009688")) |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_legend(position = "bottom")

# Crew size vs catch performance
if(nrow(crew_summary) > 0) {
  highchart() |>
    hc_add_series(
      data = crew_summary,
      type = "scatter",
      hcaes(x = boat_crew, y = avg_catch, size = records),
      name = "Crew Performance"
    ) |>
    hc_title(
      text = "Crew Size vs Average Catch Performance",
      style = list(fontSize = "16px", fontWeight = "bold")
    ) |>
    hc_subtitle(text = "Bubble size represents number of records") |>
    hc_xAxis(title = list(text = "Boat Crew Size")) |>
    hc_yAxis(title = list(text = "Average Catch Weight (kg)")) |>
    hc_colors("#673AB7") |>
    hc_tooltip(
      pointFormat = "<b>Crew Size:</b> {point.x}<br><b>Avg Catch:</b> {point.y:.2f} kg<br><b>Records:</b> {point.z}"
    ) |>
    hc_exporting(enabled = TRUE)
}
```

```{r methods-table}
trip_type_summary |>
  DT::datatable(
    caption = "Fishing Methods Comparison in Chale BMU",
    options = list(pageLength = 5, scrollX = TRUE),
    rownames = FALSE
  ) |>
  DT::formatRound(c("avg_duration", "avg_catch"), 2) |>
  DT::formatStyle(
    "percentage",
    background = styleColorBar(trip_type_summary$percentage, "#E3F2FD"),
    backgroundSize = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Species Composition & Catch Analysis

The species analysis below presents two complementary perspectives: **frequency of catch** (how often species are caught) versus **total weight contribution** (biomass significance). These metrics can show different patterns - for example, a species might be caught frequently but in small quantities, while another might be caught less often but contribute significantly to total biomass.

```{r species-analysis}
# Species summary
species_summary <- chale_bmu |>
  filter(!is.na(catch_taxon) & !is.na(catch_weight)) |>
  group_by(catch_taxon) |>
  summarise(
    records = n(),
    total_weight = sum(catch_weight, na.rm = TRUE),
    avg_weight = mean(catch_weight, na.rm = TRUE),
    median_weight = median(catch_weight, na.rm = TRUE),
    percentage = round(n() / nrow(filter(chale_bmu, !is.na(catch_taxon))) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(desc(records))

# Top 15 species by records
top_species <- species_summary |>
  slice_head(n = 15)

# Species frequency chart
top_species |>
  arrange(records) |>
  apex(
    type = "bar",
    mapping = aes(x = catch_taxon, y = records)
  ) |>
  ax_title(
    text = "Most Frequently Caught Species in Chale BMU",
    subtitle = list(text = "Ranked by frequency (number of catch records)"),
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors("#8BC34A") |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_plotOptions(
    bar = list(
      horizontal = TRUE,
      borderRadius = 4
    )
  ) |>
  ax_yaxis(title = list(text = "Number of Records"))

# Species by total weight treemap
species_treemap_data <- top_species |>
  mutate(
    value = total_weight,
    name = catch_taxon,
    colorValue = total_weight
  ) |>
  select(name, value, colorValue)

highchart() |>
  hc_add_series(
    data = species_treemap_data,
    type = "treemap",
    name = "Total Weight"
  ) |>
  hc_title(
    text = "Species by Total Catch Weight in Chale BMU",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  hc_subtitle(text = "Size represents total weight contribution") |>
  hc_colorAxis(
    minColor = "#E8F5E8",
    maxColor = "#388E3C"
  ) |>
  hc_tooltip(
    pointFormat = "<b>{point.name}</b><br>Total Weight: {point.value:.1f} kg"
  ) |>
  hc_exporting(enabled = TRUE)
```

```{r species-table}
# Species summary table
top_species |>
  mutate(
    avg_weight = round(avg_weight, 2),
    median_weight = round(median_weight, 2),
    total_weight = round(total_weight, 1)
  ) |>
  DT::datatable(
    caption = "Top Species in Chale BMU",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) |>
  DT::formatRound(c("avg_weight", "median_weight", "total_weight"), 2) |>
  DT::formatStyle(
    "percentage",
    background = styleColorBar(top_species$percentage, "#E8F4FD"),
    backgroundSize = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Catch Performance by Fishing Method

```{r catch-by-method}
# Compare catch performance between shore and boat fishing
method_catch_analysis <- chale_bmu |>
  filter(!is.na(trip_type) & !is.na(catch_weight) & catch_weight > 0) |>
  group_by(trip_type) |>
  summarise(
    total_records = n(),
    unique_trips = n_distinct(trip_id),
    total_catch = sum(catch_weight, na.rm = TRUE),
    avg_catch_per_record = mean(catch_weight, na.rm = TRUE),
    median_catch = median(catch_weight, na.rm = TRUE),
    avg_trip_duration = mean(trip_duration_hrs, na.rm = TRUE),
    catch_per_hour = total_catch / sum(trip_duration_hrs, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    catch_efficiency = total_catch / unique_trips,
    across(where(is.numeric), ~round(.x, 2))
  )

# Species composition by fishing method
species_by_method <- chale_bmu |>
  filter(!is.na(trip_type) & !is.na(catch_taxon) & !is.na(catch_weight)) |>
  group_by(trip_type, catch_taxon) |>
  summarise(
    records = n(),
    total_weight = sum(catch_weight, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(trip_type) |>
  slice_max(records, n = 8) |>
  ungroup()

# Shore vs boat catch performance comparison
method_catch_analysis |>
  pivot_longer(cols = c(avg_catch_per_record, catch_per_hour, catch_efficiency), 
               names_to = "metric", 
               values_to = "value") |>
  mutate(
    metric_label = case_when(
      metric == "avg_catch_per_record" ~ "Avg Catch per Record (kg)",
      metric == "catch_per_hour" ~ "Catch per Hour (kg/hr)",
      metric == "catch_efficiency" ~ "Catch per Trip (kg)"
    )
  ) |>
  apex(
    type = "column",
    mapping = aes(x = trip_type, y = value, fill = metric_label)
  ) |>
  ax_title(
    text = "Catch Performance: Shore vs Boat Fishing",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors(c("#FF6B35", "#00BCD4", "#4CAF50")) |>
  ax_chart(toolbar = list(show = TRUE)) |>
  ax_xaxis(title = list(text = "Fishing Method")) |>
  ax_yaxis(title = list(text = "Performance Metric")) |>
  ax_legend(position = "top")

# Top species by fishing method - Stacked bar chart
species_method_summary <- species_by_method |>
  group_by(catch_taxon) |>
  summarise(
    shore_records = sum(records[trip_type == "shore"], na.rm = TRUE),
    boat_records = sum(records[trip_type == "boat"], na.rm = TRUE),
    total_records = shore_records + boat_records,
    .groups = "drop"
  ) |>
  filter(total_records >= 20) |>
  arrange(desc(total_records)) |>
  slice_head(n = 10) |>
  pivot_longer(cols = c(shore_records, boat_records), 
               names_to = "method", 
               values_to = "records") |>
  mutate(method = str_replace(method, "_records", ""))

species_method_summary |>
  apex(
    type = "bar",
    mapping = aes(x = catch_taxon, y = records, fill = method)
  ) |>
  ax_title(
    text = "Top Species by Fishing Method",
    subtitle = list(text = "Shore vs boat fishing preferences"),
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  ax_colors(c("#FF5722", "#2196F3")) |>
  ax_chart(
    toolbar = list(show = TRUE),
    stacked = TRUE
  ) |>
  ax_plotOptions(
    bar = list(horizontal = TRUE)
  ) |>
  ax_xaxis(title = list(text = "Number of Records")) |>
  ax_yaxis(title = list(text = "Species")) |>
  ax_legend(position = "top")
```

```{r method-comparison-table}
# Method comparison summary table
method_catch_analysis |>
  DT::datatable(
    caption = "Fishing Method Performance Comparison",
    options = list(pageLength = 5, scrollX = TRUE),
    rownames = FALSE
  ) |>
  DT::formatRound(c("avg_catch_per_record", "median_catch", "avg_trip_duration", "catch_per_hour", "catch_efficiency"), 2) |>
  DT::formatStyle(
    "catch_efficiency",
    background = styleColorBar(method_catch_analysis$catch_efficiency, "#E3F2FD"),
    backgroundSize = "100% 90%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

## Gear-Specific Duration Analysis & Data Quality

```{r outlier-detection}
# Function to detect outliers using IQR method by gear type
detect_outliers_by_gear <- function(data, duration_col, gear_col, k = 1.5) {
  data |>
    group_by({{gear_col}}) |>
    mutate(
      q1 = quantile({{duration_col}}, 0.25, na.rm = TRUE),
      q3 = quantile({{duration_col}}, 0.75, na.rm = TRUE),
      iqr = q3 - q1,
      lower = q1 - k * iqr,
      upper = q3 + k * iqr,
      is_outlier = {{duration_col}} < lower | {{duration_col}} > upper,
      # For traps and nets, use more lenient thresholds as they can operate for days
      is_outlier = case_when(
        str_detect(tolower({{gear_col}}), "trap|net|seine") ~ {{duration_col}} > (q3 + 3 * iqr) | {{duration_col}} < (q1 - 3 * iqr),
        TRUE ~ is_outlier
      )
    ) |>
    ungroup() |>
    select(-q1, -q3, -iqr, -lower, -upper)
}

# Trip duration analysis considering gear types
trip_duration_clean <- chale_bmu |>
  filter(!is.na(trip_duration_hrs) & trip_duration_hrs >= 0 & trip_duration_hrs < 200) # Increased threshold for traps

# Analyze duration patterns by gear type
gear_duration_summary <- chale_bmu |>
  filter(!is.na(trip_duration_hrs) & trip_duration_hrs >= 0 & !is.na(gear_type)) |>
  group_by(gear_type) |>
  summarise(
    records = n(),
    median_duration = median(trip_duration_hrs, na.rm = TRUE),
    q75_duration = quantile(trip_duration_hrs, 0.75, na.rm = TRUE),
    max_duration = max(trip_duration_hrs, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(median_duration))

# Detect outliers considering gear type
duration_outliers <- chale_bmu |>
  filter(!is.na(trip_duration_hrs) & trip_duration_hrs >= 0 & !is.na(gear_type)) |>
  detect_outliers_by_gear(trip_duration_hrs, gear_type) |>
  filter(is_outlier) |>
  arrange(desc(trip_duration_hrs))

# Simple outlier detection function for catch weights
detect_outliers <- function(x, k = 1.5) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - k * iqr
  upper <- q3 + k * iqr
  return(x < lower | x > upper)
}

# Catch weight outlier analysis
weight_outliers <- chale_bmu |>
  filter(!is.na(catch_weight) & catch_weight > 0) |>
  mutate(is_outlier = detect_outliers(catch_weight)) |>
  filter(is_outlier) |>
  arrange(desc(catch_weight))

# Gear duration summary visualization
if(nrow(gear_duration_summary) > 0) {
  gear_duration_summary |>
    apex(
      type = "bar",
      mapping = aes(x = gear_type, y = median_duration)
    ) |>
    ax_title(
      text = "Gear Operational Duration Patterns",
      subtitle = list(text = "Median duration by gear type (hours)"),
      style = list(fontSize = "16px", fontWeight = "bold")
    ) |>
    ax_colors("#FF5722") |>
    ax_plotOptions(
      bar = list(horizontal = TRUE)
    ) |>
    ax_xaxis(title = list(text = "Median Duration (hours)")) |>
    ax_yaxis(title = list(text = "Gear Type"))
}

# Trip duration distribution by gear type
duration_by_gear <- chale_bmu |>
  filter(!is.na(trip_duration_hrs) & trip_duration_hrs >= 0 & !is.na(gear_type) & trip_duration_hrs < 100) |>
  group_by(gear_type) |>
  filter(n() >= 10) |> # Only include gear types with sufficient data
  ungroup()

if(nrow(duration_by_gear) > 0) {
  duration_by_gear |>
    apex(
      type = "boxplot", 
      mapping = aes(x = gear_type, y = trip_duration_hrs)
    ) |>
    ax_title(
      text = "Gear Operational Duration Distribution",
      subtitle = list(text = "Duration patterns vary significantly by gear type"),
      style = list(fontSize = "16px", fontWeight = "bold")
    ) |>
    ax_colors("#FF5722") |>
    ax_xaxis(title = list(text = "Gear Type")) |>
    ax_yaxis(title = list(text = "Duration (hours)"))
}

# Catch weight distribution histogram
weight_data <- chale_bmu$catch_weight[!is.na(chale_bmu$catch_weight) & chale_bmu$catch_weight > 0]
weight_hist <- hist(log10(weight_data), breaks = 30, plot = FALSE)

highchart() |>
  hc_add_series(
    data = data.frame(
      x = weight_hist$mids,
      y = weight_hist$counts
    ),
    type = "column",
    hcaes(x = x, y = y),
    name = "Frequency"
  ) |>
  hc_title(
    text = "Catch Weight Distribution in Chale BMU",
    style = list(fontSize = "16px", fontWeight = "bold")
  ) |>
  hc_subtitle(text = "Log10 scale distribution") |>
  hc_xAxis(
    title = list(text = "Log10(Catch Weight)")
  ) |>
  hc_yAxis(title = list(text = "Frequency")) |>
  hc_colors("#00BCD4") |>
  hc_chart(zoomType = "x") |>
  hc_exporting(enabled = TRUE)
```

### Gear-Adjusted Duration Analysis

::: highlight-box
**Gear Duration Patterns:** - **Different gear types** have naturally different operational durations - **Traps/Nets:** Can operate for days (24-72+ hours is normal) - **Hooks/Lines:** Typically shorter operations (2-12 hours) - **`r nrow(duration_outliers)`** records flagged as unusual even considering gear type

**Gear-Adjusted Outliers:** - **Longest unusual duration:** `r if(nrow(duration_outliers) > 0) paste(round(max(duration_outliers$trip_duration_hrs, na.rm = TRUE), 1), "hours") else "None"` - **Analysis:** Outliers now consider gear-specific normal ranges

**Catch Weight Outliers:** - **`r nrow(weight_outliers)`** records with extreme catch weights - **Heaviest catch:** `r max(weight_outliers$catch_weight, na.rm = TRUE)` kg - **Analysis:** Large catches should be verified against typical species sizes
:::

```{r outlier-tables}
# Gear duration summary table
if(nrow(gear_duration_summary) > 0) {
  gear_duration_summary |>
    mutate(
      median_duration = round(median_duration, 1),
      q75_duration = round(q75_duration, 1),
      max_duration = round(max_duration, 1)
    ) |>
    DT::datatable(
      caption = "Gear Type Duration Patterns",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) |>
    DT::formatStyle(
      "median_duration",
      background = styleColorBar(gear_duration_summary$median_duration, "#E8F4FD"),
      backgroundSize = "100% 90%",
      backgroundRepeat = "no-repeat",
      backgroundPosition = "center"
    )
}

# Gear-adjusted duration outliers
if(nrow(duration_outliers) > 0) {
  duration_outliers |>
    select(trip_id, fisher_name, gear_type, trip_duration_hrs, catch_weight, trip_type) |>
    slice_head(n = 10) |>
    mutate(
      trip_duration_hrs = round(trip_duration_hrs, 2),
      fisher_name = str_replace(fisher_name, "_", " ")
    ) |>
    DT::datatable(
      caption = "Gear-Adjusted Duration Outliers",
      options = list(pageLength = 5, scrollX = TRUE),
      rownames = FALSE
    )
}

# Top weight outliers  
if(nrow(weight_outliers) > 0) {
  weight_outliers |>
    select(trip_id, fisher_name, catch_taxon, catch_weight, trip_type) |>
    slice_head(n = 10) |>
    mutate(fisher_name = str_replace(fisher_name, "_", " ")) |>
    DT::datatable(
      caption = "Top 10 Catch Weight Outliers in Chale BMU", 
      options = list(pageLength = 5, scrollX = TRUE),
      rownames = FALSE
    )
}
```

## Data Completeness Assessment

```{r data-quality}
# Missing data analysis for Chale BMU
missing_summary <- 
  chale_bmu |>
    dplyr::mutate(boat_type = dplyr::case_when(trip_type == "shore" & is.na(boat_type) ~ "foot", TRUE ~ boat_type)) |> 
  summarise_all(~sum(is.na(.))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") |>
  mutate(
    missing_percent = round(missing_count / nrow(chale_bmu) * 100, 2),
    data_quality = case_when(
      missing_percent == 0 ~ "Excellent",
      missing_percent < 5 ~ "Good", 
      missing_percent < 20 ~ "Fair",
      TRUE ~ "Poor"
    )
  ) |>
  arrange(desc(missing_percent))

# Data quality visualization
missing_chart_data <- missing_summary |>
  filter(missing_count > 0) |>
  arrange(desc(missing_percent)) |>
  mutate(
    color = case_when(
      data_quality == "Excellent" ~ "#4CAF50",
      data_quality == "Good" ~ "#FF9800", 
      data_quality == "Fair" ~ "#FF5722",
      data_quality == "Poor" ~ "#9C27B0"
    )
  )

if(nrow(missing_chart_data) > 0) {
  highchart() |>
    hc_add_series(
      data = missing_chart_data,
      type = "bar",
      hcaes(x = variable, y = missing_percent, color = color),
      name = "Missing Data %"
    ) |>
    hc_title(
      text = "Data Completeness in Chale BMU",
      style = list(fontSize = "16px", fontWeight = "bold")
    ) |>
    hc_subtitle(text = "Percentage of missing values by variable") |>
    hc_xAxis(
      title = list(text = "Variable"),
      categories = missing_chart_data$variable
    ) |>
    hc_yAxis(
      title = list(text = "Missing Data (%)"),
      min = 0
    ) |>
    hc_tooltip(
      pointFormat = "<b>{point.category}</b><br>Missing: {point.y:.1f}%"
    ) |>
    hc_legend(enabled = FALSE) |>
    hc_exporting(enabled = TRUE)
}
```

::: card
### Data Quality Summary for Chale BMU

```{r quality-table}
missing_summary |>
  DT::datatable(
    caption = "Missing Data Analysis",
    options = list(pageLength = 10),
    rownames = FALSE
  ) |>
  DT::formatStyle(
    "data_quality",
    backgroundColor = styleEqual(
      c("Excellent", "Good", "Fair", "Poor"),
      c("#d4edda", "#fff3cd", "#f8d7da", "#e2e3f4")
    )
  )
```
:::

## Key Insights & Recommendations

::: highlight-box
### 🎣 **Chale BMU Fishing Characteristics**

-   **Multi-Site Activity:** Three landing sites: Chale-Jeza (45%), Chale Island (34.5%), Kinondo Ngwani (20.6%)
-   **Method Variations:** Chale Island almost exclusively shore-based (99.3%), while Chale-Jeza (71% shore) and Kinondo Ngwani (65.4% shore) have more mixed approaches
-   **Peak Activity:** Early morning departures (4-8 AM) account for 82% of all BMU trips
-   **Gender Participation:** Overwhelmingly male-dominated (98.7%) with minimal female participation
-   **Fisher Population:** 693 unique fishers across the BMU with 443 at Chale-Jeza alone

### 🏝️ **Inter-Site Variations**

-   **Activity Concentration:** Chale-Jeza dominates with 45% of all records, nearly twice Chale Island's activity
-   **Method Specialization:** Chale Island is almost purely shore-based, contrasting with mixed methods at other sites
-   **Fisher Density:** Highest fisher numbers at Chale-Jeza (443), moderate at Chale Island (251), lowest at Kinondo Ngwani (89)
-   **Trip Patterns:** Different trip-to-record ratios suggest varying catch logging practices between sites

### 🐟 **BMU-Wide Species Composition**

-   **Dual Species Dynamics:** Octopus dominates by frequency (41.2% of records) while marbled parrotfish dominates by biomass (35,898 kg total weight)
-   **Fishing Strategy Patterns:** Frequent small octopus catches vs less frequent but larger parrotfish catches reflect different targeting strategies
-   **Biomass vs Activity:** Octopus shows highest fishing activity but marbled parrotfish contributes most to total catch weight
-   **Diverse Portfolio:** "Other" category (12.4%) and Rabbitfish (10.5%) show significant catch diversity
-   **Reef Ecosystem Health:** Multiple parrotfish and surgeonfish species indicate healthy reef fish targeting across BMU

### 📊 **Data Quality & Coverage**

-   **Multi-Site Coverage:** Comprehensive monitoring across BMU landing sites
-   **Temporal Consistency:** Reliable timing data across all sites
-   **Some Anomalies:** Extreme values requiring validation across sites
-   **Good Completeness:** Strong data coverage for most core variables

### 💡 **BMU-Level Recommendations**

1.  **Implement BMU-wide management** considering inter-site variations
2.  **Site-specific monitoring** tailored to each landing site's characteristics\
3.  **Cross-site fisher mobility** analysis to understand movement patterns
4.  **Coordinated data validation** for extreme values across all sites
5.  **BMU-wide sustainability** monitoring for key species like octopus
6.  **Gender inclusion initiatives** across all landing sites
:::

------------------------------------------------------------------------

*Report generated on `r Sys.Date()` using Quarto. Data source: ABALOBI Monitor App, Chale Beach Management Unit (BMU) analysis.*